Data Descriptor for Nature Scientific Data

### 2024-10-30 manuscript/manuscript-nature/manuscript-nature.qmd
res-part-ii

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America. We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.

The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.

@@

We first focus on analysing the variability in COVID-19's impact on daily mobility levels across different population groups. Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022. These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset. May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures. March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis. Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index. The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$. Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction. These changes show remarkably consistent patterns across countries for both types of classification.

@@

![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig-impact.png)

@@

Next we focus on modelling intergroup differences in mobility trends over time towards baseline levels of mobility, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups. Figure @Fig-recovery-density and @Fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively. Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels. The background of each plot in Figures @Fig-recovery-density and @Fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns. This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.

A visual inspection of the evolution patterns by population density category, as shown in @Fig-recovery-density, reveals two main commonalities across the three countries. First, more densely populated areas consistently record lower levels of mobility throughout the study period. Second, there is a general upward trend in mobility across all population density categories for all countries. However, variations exist in the patterns for each country. While Chile experiences greater overall losses in mobility compared to Argentina, certain areas in Argentina reach levels comparable to the pre-pandemic baseline. In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses.

![](figures/Fig-recovery-density.png)

Based on a visual inspection of @Fig-recovery-rdi, the evolution patterns by relative deprivation index category mirror those observed by population density category. First, less deprived areas consistently record lower levels of mobility throughout the study period. In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline. However, these significant relative changes do not necessarily correspond to large population numbers. In Colombia, for example, more deprived regions are often low-density areas that typically contain small populations. As a result, minor fluctuations in mobility can lead to substantial percentage changes. Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.

![](figures/Fig-recovery-rdi.png)
Next, we quantitatively model the patterns observed in @Fig-recovery-density and @Fig-recovery-rdi

![](figures/Fig-re-density.pdf)

![](figures/Fig-re-rdi.pdf)

>>>>

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their relative level of socioeconomic deprivation.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.

@@

We first focus on analysing the variability in COVID-19's impact on daily mobility levels across different population groups.
@fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022.
These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset.
May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures.
March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis.
@fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index.
The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$.
Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction.
These changes show remarkably consistent patterns across countries for both types of classification.

@@

![Changes in the intensity of flows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis at the origin location of each flow.](figures/Fig-impact.pdf){#fig-impact}

@@

Next we focus on modelling intergroup differences in mobility trends over time, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
@fig-recovery-density and @fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
The background of each plot in Figures @fig-recovery-density and @fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.

A visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density, reveals two main commonalities across the three countries.
First, there is a general upward trend in mobility across all population density categories for all countries.
Second, more densely populated areas consistently record lower levels of mobility throughout the study period.
However, there exist variations in the patterns by country.
Chile experiences greater overall losses in mobility compared to Argentina, while certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses. Generally, the termporal patterns display fluctuations from the general trend which are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.

![](figures/Fig-recovery-density.png){#fig-recovery-density}

The evolution patterns by relative deprivation index category mirror those observed by population density category, as it can be deduced from @Fig-recovery-rdi based on visual inspection. 
First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
Second, less deprived areas consistently record lower levels of mobility throughout the study period.
In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline.
However, these significant relative changes do not necessarily correspond to large population numbers.
In Colombia, for example, more deprived regions are often low-density areas that typically contain small populations.
As a result, minor fluctuations in mobility can lead to substantial percentage changes.
Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.

![](figures/Fig-recovery-rdi.png){#fig-recovery-rdi}

Next, we quantitatively model the patterns observed in @fig-recovery-density and @fig-recovery-rdi

![](figures/Fig-re-density.pdf){#fig-re-density}

![](figures/Fig-re-rdi.pdf){#fig-re-rdi}

![](figures/Fig-re-rdi.pdf){#Fig-re-rdi}
